A Bayesian Framework for Learning Regular Expressions

### 2012-12-09 paper/paper.tex
Corrections.

In doing so, we first generated a Bayesian model that, given a set of input strings $S$, generated a regular expression that describes $S$ with the highest posterior probability. This model is discussed in the first part of this paper. We then ran a behavioral experiment in which subjects were asked whether or not a given string $s$ fits with a set of strings $S$. We compare the result of that experiment with results generated given an extension of our model.

@@

Below, we define the prior probability such that our model biases smaller hypotheses, and the likelihood of generating a given hypothesis is inversely proportional to its cost. Our model aims to balance generality with specificity. For example, given `abc, abd', we wish to favor the regex \verb!ab[c|d]! over \verb!A*! where \verb!A! represents any letter,\footnote{See the Appendix for a brief summary of our regex notation.} which is less likely given the size principle. At the same time, given `aba', `abc', `abd', `abe', `abf', we may wish to favor \verb!abA! over \verb!ab[a|c|d|e|f]!.

@@

Regular expressions are equivalent to deterministic finite automata (DFA). Because DFA are often easier to reason with, we use them to represent regular expressions in our model (See Figure~\ref{base_hypothesis} for an example DFA.)

Further, we define the probability with which a DFA would stop generating a string at a start state . For instance, given the regular expression \verb!a*!, we would need to define the probability of generating strings `', `a', `aa', `aa', etc.

Our generative version of the DFA is not to be confused with a probabilistic atumaton, which has transition probabilities much as described above and a final acceptance threshold. Our generative automata does not have this threshold, and instead has an added parameter which controls the likelihood of stopping to generate a string.

As our alphabet, we use the set of all lowercase letters. This set, we feel, has an intuitive boundary, as subjects seeing strings of all lowercase letters are unlikely to suppose an arbitrary uppercase letter or special character, and thus we assume the expressiveness of wildcards to be fixed.

@@

\subsection{Generating Hypotheses}
The hypothesis space is initiated with the base hypothesis, the term applied to a DFA that accepts strictly the input strings. We begin generating a series of consecutive states, where the transition between the first state and the second state is governed by the first letter of the first string, the transition between the second and the third is governend by the the second letter, and so on. For each additional string, if the first i characters overlap with those of any of the first i characters of one previous string but differ in the $i+1$th, we add no new states to the DFA until the transition between the i-1th state and the ith state. We then generate new states for the remaining characters of the string. See Figure~\ref{base_hypothesis} for an example of the first hypothesis generated from a set $S = $\{`ababb', `abbab'\}.
  \begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.4]{base_hypothesis.png}
\end{center}
\caption{An example base hypothesis DFA generated from the strings `ababb' and `abbab'. Note that the two strings share the first two states, but that the fifth and tenth are distinct.} 
\label{base_hypothesis}
\end{figure}

Next, hypotheses are added to the hypothesis space through two processes of generalization: merging and applying wildcards.

\subsubsection{Merging} Merging is the process of combining states of a DFA. The newly create state has all incoming and outgoing transitions of its predecessors. Thus, the DFA will continue to accept the same set of strings as the original DFA. However, the DFA will often accept more strings after merging, thus merging states results in a more generalized DFA.

If merging is continued, the DFA will eventually have only one state with all transitions leading to itself. Such a state accepts all strings containing only zero or more of the original characters. A series of merges are illustrated in Figure~\ref{merge-figure}.

@@

\subsubsection{Enforcing Determinism} Merging states and replacing transitions with wildcards can allow one state to have non-unique outbound transitions. However, this turns our representation into a nondeterministic finite automaton (NFA) instead of a DFA. Both are equivalent, but the latter afford easier manipulation for our experiment. Thus, to convert the NFA back to a DFA, we use the subset construction, the effect of which is illustrated in both Figure~\ref{merge-figure} and Figure~\ref{wildcard-figure}.

\subsection{Searching the Hypothesis Space} The number of hypotheses grows quickly with the number of states in the base hypothesis. As a lower bound, the number of hypotheses is at least as large as the number of possible starting nodes. Because the starting node could be merged with any subset of the remaining $n-1$ nodes in the DFA, there are $2^{n-1}=\Omega(2^n)$ possible start states. Thus, the hypothesis space is at least exponential in the number of starting nodes in the regular expression.

@@

An upper bound for the complexity of model merging is much more extreme. Performing subset construction for an NFA with $n$ states can produce, in the worst case, a DFA with $2^n$ states. Assuming that any subset of these states also produces a valid DFA, we can have at most $2^{2^n-1}$ DFA. \footnote{See the Appendix for detailed sizing of the hypothesis space.}

We resolve this issue by performing beam search. In beam search, we perform breadth first search, but for any given depth of the search tree, we only consider children of the $m$ most promising hypotheses. In our case, the most promising hypotheses were defined as those with the highest posterior probability. (For a more detailed discussion of Beam Search, see \cite{HwangStuhlmullerGoodman2011}.

\section{Inference with DFA}
The posterior probability for a DFA as a hypothesis is given by the following.
\begin{align*}
	p(h|S) \propto p(S|h)p(h)
\end{align*}
Thus, inferring the best DFA given a set of input strings was essentially a problem in balancing simplicity with conservation of belief. That is, DFA were biased a priori for being simpler, but simpler DFA tended to accept more strings.

\subsection{Prior Probability}
We assigned the prior probability to the hypotheses (i.e. regexes) based on the size of the DFA and the number of transitions. In the following equation, $|S|$ is the number of states in the DFA and $|\delta|$ is the number of tranisitons in the DFA, where a wildcard counts as only one transition. 
\begin{align*}
	P(h) \propto e^{-\alpha|S| - \beta|\delta|}
\end{align*}
Thus, we have biased smaller DFA that use wildcards instead of literal characters. [cite Program Merging model]

@@

Let the string $s$ be the concatenation of characters $\{a_0,a_1,...,a_{n-1}\}$. We define the probability of generating a string as the product of the probability of generating the letters in sequence $p(a_0,a_1,...,a_{n-1})$ and the product of stopping the string generation on an accept state $p(\text{stop}|h)$.
\begin{align*}
	p(s|h) = p(a_0,a_1,...,a_{n-1}|h)p(\text{stop}|h)
\end{align*}

@@

To simplify inference, we assume uniform probability over all possible outbound transitions for a given state. Formally, the probability of making a state transition is given by:
\begin{align*}
	p(\delta_h(r_i, a_i)) &= \frac{1}{\sum_{a\in A} 1\{\delta_h(r_i, a)\in Q_h\}}
\end{align*}
where $Q_h$ is the set of states in hypothesis $h$ and $1\{\cdot\}$ is the indicator function, which evaluates to $1$ if the enclosed proposition is true and $0$ otherwise. $\delta_h(r_i, a)\in Q_h$ indicates that the state transition leads to another state of the DFA, so the denominator counts the number of characters for which a transition is defined. In the case of the wildcard `A', the transition is defined for all 26 letters, so the denominator is 26.

For the cases in which a string may alternatively be accepted or continue, one must assign a probability for accepting or continuing the string. For instance, consider the DFA in part (c) of Figure~\ref{merge-figure}. The DFA may generate the empty string, `', or it may generate any arbitrary string with `a' and `b', like `ababbba'. We define the probability of stopping at a given string to be a fixed value $\tau$. This makes shorter strings more likely to be generated.

@@

Our equation for $p(\text{stop}|h)$ is thus:
\begin{align*}
	p(\text{stop}|h) &= \left\{
		\begin{array}{rl}
		\frac{1}{b^2+3b+1} & r_n \in F \wedge \exists a : \delta(r_n, a)
		\\ \frac{1}{b+1} & r_n \in F \wedge \not \exists a : \delta(r_n, a)
		\\ 0 & \text{otherwise}
		\end{array}
		\right.
\end{align*}

\section{Experiment Design and Model Fitting}
Having generated the above Bayesian model for selection of the regular expression with the highest posterior probability, we then conducted a behavioral experiment.  41 subjects (Median age 34, std-dev 13.3) were surveyed on Amazon Mechanical Turk. Two groups (one of 20, one of 21) were presented with 10 yes or no questions in which they were asked to determine whether a given string ``fit'' with a list of strings. A sample question was as follows:
\begin{center}Given the list:

@@

The experimental data served two purposes. Using the responses to four out of 23 questions, we set the parameters for alpha and beta. We then used the remaining questions to test our model's performance in a regex game as shown below.

@@

To fit the $\beta$ term, we observed experiments that would indicate the transition point between a preference for generality over specificity for wildcard-based experiments (e.g. the first two experiments outlined in Table~\ref{parameter-table}). Utilizing the fact that increasing the ratio of repeated strings to new strings decreases the likelihood of a wildcard, and decreasing the length of the sample strings would increases the likelihood of a wildcard, the first two experiments suggest a decision point. That is, if the DFA model is a valid representation of the subjects' responses, then the $\beta$ term should be between $3$ and $5$. For further experimentation, we fixed it at $4$.

@@

\subsection{The Regex Game}
Given the fitted model, we then considered the following task:

Given a language $\mathcal{L}$, a set of regular expressions $\mathcal{H}$ (where these are candidates for $\mathcal{L}$), a set of strings drawn $S$ from $\mathcal{L}$, and a string $s$, we define

\begin{align*}
	p(s \in \mathcal{L} | S \subseteq {\mathcal{L}}) = \sum_{h \in \mathcal{H}} 1 \{s \in h\} p(h|S)
\end{align*}

For human participants, because the hypothesis is unobserved, we take $p(s \in \mathcal{L} | S \subseteq {\mathcal{L}}) $ to be the average acceptance of a test string given a set $S$.

@@

Observe that the model's performence is effectively all or none. In twelve of the trials, the model's predictions for  $p(s \in \mathcal{L} | S \subseteq {L}) $ was either 0 or 1. This suggests that the posterior is dominated by hypotheses with high likelihoods, all of which predict the same result.

We might explain the lack of diversity in the hypothesis set by

We note that in cases 16 (S = \{`yyyt', `yt', `yyyt', `yyyt', `yt'\}, s= `yyyyyt'), 17 (S = \{`jojojojo', `jojojojo', `jojojojo'\}, s = `jojo'), and 18 (S = \{`twtwtw', `twtwtw', `twtwtw'\}, s =  `tw'), it is clear that the model weighs a smaller number of transitions greater than human subjects do. However, because this same parameter in part controls for wildcards, lowering it reduces performance on many other cases.

>>>>

In doing so, we first generated a Bayesian model that, given a set of input strings $S$, generated a regular expression that describes $S$ with the highest posterior probability. This model is discussed in the first part of this paper. We then discuss a behavioral experiment in which subjects were asked whether or not a given string $s$ fits with a set of strings $S$. Finally, we compare the result of that experiment with results generated given an extension of our model.

@@

Below, we define the prior probability of a hypothesis $h$ and the likelihood of generating a string $s$. Our model aims to balance generality with specificity. For example, given `abc' and `abd', we wish to favor the regex \verb!ab[c|d]! over \verb!A*! (where \verb!A! represents any letter\footnote{See the Appendix for a brief summary of our regex notation.}) which is less likely given the size principle. At the same time, given `aba', `abc', `abd', `abe', `abf', we may wish to favor \verb!abA! over \verb!ab[a|c|d|e|f]!.

@@

Regular expressions are equivalent to deterministic finite automata (DFA). Because DFA are often easier to reason with, we use them to represent regular expressions in our model. See Figure~\ref{base_hypothesis} for an example DFA.

To define the likelihood of a string given a DFA, we view the DFA as generative. To do so, we attribute transition probabilities. For instance, the regular expression \verb!a[b|c]! may have equal likelihood of generating `ab' or `ac'. Further, we define the probability with which a DFA would stop generating a string at an accept state. For instance, given the regular expression \verb!a*!, we would need to define the probability of generating strings `', `a', `aa', `aa', etc.

Our generative version of the DFA is not to be confused with a probabilistic automaton, which has transition probabilities much as described above and a final acceptance threshold. Our generative automata does not have this threshold, and also has an added parameter which controls the likelihood of stopping to generate a string.

As our alphabet, we use the set of all lowercase letters. This set, we feel, has an intuitive boundary, as subjects seeing strings of all lowercase letters are unlikely to suppose an arbitrary uppercase letter or special character. Thus, we can say that the wildcard replaces any of the 26 possible transitions.

@@

\subsection{Generating Hypotheses}
We initialize the hypothesis with the \emph{base hypothesis}, the DFA that accepts strictly the input strings. Before any strings are added to the base hypothesis, the DFA is simply a start state. Next, given an input string, for each character of the input string, the model adds a new state and a transition. When adding a string to the model, wherever the state transition is already defined, a new state is not created. This has the effect of merging paths that have similar starting transitions. See Figure~\ref{base_hypothesis} for an example of the first hypothesis generated from a set $S = $\{`ababb', `abbab'\}.

\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.4]{base_hypothesis.png}
\end{center}
\caption{An example base hypothesis DFA generated from the strings `ababb' and `abbab'. Note that the two strings share the first two states, but that the fifth and tenth are distinct.} 
\label{base_hypothesis}
\end{figure}

Next, we iteratively generalize hypotheses in our hypothesis space, starting with the base hypothesis. New hypotheses are generated through two processes of generalization: merging and applying wildcards.

\subsubsection{Merging} Merging is the process of replacing two states of a DFA with one state. The newly created state has all incoming and outgoing transitions of its predecessors. Thus, the DFA will continue to accept the same set of strings as the original DFA. However, the DFA will often accept more strings after merging, thus merging results in a more generalized DFA.

If merging is continued, the DFA will eventually have only one state with all transitions leading to itself. Such a state accepts all strings containing zero or more of the original characters. A series of merges are illustrated in Figure~\ref{merge-figure}.

@@

\subsubsection{Enforcing Determinism} Merging states and replacing transitions with wildcards can allow one state to have non-unique outbound transitions. This turns our representation into a nondeterministic finite automaton (NFA) instead of a DFA. Both are equivalent, but the latter is an inconsistent representation for our model. Thus, to convert the NFA back to a DFA, we use the subset construction, the effect of which is illustrated in both Figure~\ref{merge-figure} and Figure~\ref{wildcard-figure}.

\subsection{Searching the Hypothesis Space} The number of hypotheses grows quickly with the number of states in the base hypothesis. As a lower bound, the number of hypotheses is at least as large as the number of possible starting nodes. Because the starting node could be merged with any subset of the remaining $n-1$ nodes in the DFA, there are $2^{n-1}=\Theta(2^n)$ possible start states. Thus, the hypothesis space is at least exponential in the number of starting nodes in the regular expression.

@@

An upper bound for the complexity of model merging is much more extreme. Performing subset construction for an NFA with $n$ states can produce, in the worst case, a DFA with $2^n$ states. Assuming that any subset of these states also produces a valid DFA, we can have at most $2^{2^n-1}$ DFA. This, however, is a very conservative estimate.\footnote{See the Appendix for detailed sizing of the hypothesis space.}

We resolve this issue by performing beam search. In beam search, we perform breadth first search, but for any given depth of the search tree, we only consider children of the $m$ most promising hypotheses. In our case, the most promising hypotheses were defined as those with the highest posterior probability. For a more detailed discussion of Beam Search, see \citeA{HwangStuhlmullerGoodman2011}.

\section{Inference with DFA}
The posterior probability for a DFA as a hypothesis is given by the following.
\begin{align*}
	p(h|S) \propto p(S|h)p(h)
\end{align*}
Thus, inferring the best DFA given a set of input strings was essentially a problem in balancing simplicity with conservation of belief. That is, DFA were biased a priori for being simpler, but simpler DFA tended to accept more strings and were accordingly penalized.

\subsection{Prior Probability}
We assigned the prior probability to the hypotheses (i.e. regexes) based on the size of the DFA and the number of transitions, similar to \citeA{HwangStuhlmullerGoodman2011}. In the following equation, $|S|$ is the number of states in the DFA and $|\delta|$ is the number of tranisitons in the DFA, where a wildcard counts as only one transition. 
\begin{align*}
	P(h) \propto e^{-\alpha|S| - \beta|\delta|}
\end{align*}
Thus, we have biased smaller DFA that use wildcards instead of literal characters.

@@

Let the string $s$ be the concatenation of characters $\{a_0,a_1,...,a_{n-1}\}$. We define the probability of generating a string as the product of the probability of generating the letters in sequence $p(a_0,a_1,...,a_{n-1})$ and the probability of stopping the string generation on an accept state $p(\text{stop}|h)$.
\begin{align*}
	p(s|h) = p(a_0,a_1,...,a_{n-1}|h)p(\text{stop}|h)
\end{align*}

@@

To simplify inference, we assume uniform probability over all possible outbound transitions for a given state. Formally, the probability of making a state transition is given by:
\begin{align*}
	p(\delta_h(r_i, a_i)) &= \frac{1}{\sum_{a\in A} 1\{\delta_h(r_i, a)\in Q_h\}}
\end{align*}
where $Q_h$ is the set of states in hypothesis $h$ and $1\{\cdot\}$ is the indicator function, which evaluates to $1$ if the enclosed proposition is true and $0$ otherwise. $\delta_h(r_i, a)\in Q_h$ indicates that the state transition leads to another state of the DFA, so the denominator counts the number of characters for which a transition is defined. In the case of the wildcard \verb!A!, the transition is defined for all 26 letters, so the denominator is 26.

For the cases in which a DFA may output a string or continue extending it, one must assign a probability for each option. For instance, consider the DFA in part (c) of Figure~\ref{merge-figure}. The DFA may generate the empty string, `', or it may generate any arbitrary string with `a' and `b', like `ababbba'. We define the probability of stopping at a given string to be a fixed value $\tau$. This makes shorter strings more likely to be generated.

@@

Our equation for $p(\text{stop}|h)$ is thus:
\begin{align*}
	p(\text{stop}|h) &= \left\{
		\begin{array}{rl}
		\frac{1}{b^2+3b+1} & r_n \in F_h \wedge \exists a : \delta_h(r_n, a) \in Q_h
		\\ \frac{1}{b+1} & r_n \in F_h \wedge \not \exists a : \delta_h(r_n, a) \in Q_h
		\\ 0 & \text{otherwise}
		\end{array}
		\right.
\end{align*}

\section{Experimental Design and Model Fitting}
Having generated the above Bayesian model for selection of the regular expression with the highest posterior probability, we then conducted a behavioral experiment.  41 subjects (Median age 34, std-dev 13.3) were surveyed on Amazon Mechanical Turk. Two groups (one of 20, one of 21) were presented with 10 yes or no questions in which they were asked to determine whether a given string ``fit'' with a list of strings. A sample question was as follows:
\begin{center}Given the list:

@@

The experimental data served two purposes. Using the responses to four out of 23 questions, we set the values for $\alpha$ and $\beta$. We then used the remaining questions to test our model's performance in a regex game as shown below.

@@

To fit the $\beta$ term, we observed experiments that would indicate the transition point between a preference for generality over specificity for wildcard-based experiments (e.g. the first two experiments outlined in Table~\ref{parameter-table}). Utilizing the fact that increasing the number of repeated strings decreases the likelihood of a wildcard, and decreasing the length of the sample strings would increases the likelihood of a wildcard, the first two experiments suggest a decision point. That is, the first experiment suggests that to be in agreement with 70\% of the subjects, the model should have a $\beta$ greater 5. The second experiment suggests that to be in agreement with 55\% of the subjects, the model should have a $\beta$ less than 3. So, the $\beta$ term should be between $3$ and $5$. For further experimentation, we fixed it at $4$.

@@

\subsection{The Regex Game}
We then considered the following task:

Given a language $\mathcal{L}$, a set of regular expressions $\mathcal{H}$ that recognize a superset of $\mathcal{L}$, a set of strings drawn $S$ from $\mathcal{L}$, and a string $s$, we define
\begin{align*}
	p(s \in \mathcal{L} | S \subseteq {\mathcal{L}}) = \sum_{h \in \mathcal{H}} 1 \{s \in h\} p(h|S)
\end{align*}

For human participants, because the hypothesis is unobserved, we take $p(s \in \mathcal{L} | S \subseteq {\mathcal{L}}) $ to be the average acceptance of a test string given a set $S$. This would have the analogous effect as assuming that the subjects perform probability matching.

@@

Observe that the model's performence is effectively all or none. In twelve of the trials, the model's predictions for  $p(s \in \mathcal{L} | S \subseteq {L}) $ was either 0 or 1. After performing inference, the model generally included one or two hypotheses amounting to over 90\% of the voting weight. Thus, though our inference process considered many hypotheses at once, only a very small number had significant impact on the result.

We note that in cases 16 (S = \{`yyyt', `yt', `yyyt', `yyyt', `yt'\}, s= `yyyyyt'), 17 (S = \{`jojojojo', `jojojojo', `jojojojo'\}, s = `jojo'), and 18 (S = \{`twtwtw', `twtwtw', `twtwtw'\}, s =  `tw'), it is clear that the model more strongly prefers the use of the Kleene star. However, because this same parameter in part controls for wildcards, lowering it reduces performance on many other cases.
